decompr: GVC decomposition in R

### 2015-01-29 draft.tex
update implementation and introduction

@@

\begin{equation}
VB = V (I-A)^{-1},
\end{equation}
where \textit{V} is a N x N matrix with the diagonal representing the direct value added contribution of the industries,
\textit{A} is the Input-Output coefficient matrix with dimension N x N, i.e. it gives the direct input flows between
industries required for 1\$ of output, $B = (I-A)^{-1}$ is then the so called Leontief inverse, and \textit{T} indicates
a matrix transpose operation. \textit{VB} gives thus a N x N matrix of so called value added multipliers,
which denote the amount of value added that the production of an industry's 1\$ of output or exports brings about
in all other industries. Looking from the perspective of the supplying industries, the matrix gives the value added that
they contribute to the using industry's production. If we multiply it with a N x N matrix whose diagonal specifies each
industry's total output, we get value added origins as absolute values instead of shares.

The application of the Leontief insight to ICIOs as opposed to national Input-Output tables for our source decomposition
is straightforward. \textit{V} refers now to a vector of direct value added contribution of all industries across
the different countries. Its dimension is correspondingly 1 x GN. \textit{A} is now of dimension GN x GN
and gives the industry flows including cross border relationships. Since we are interested in the value added origins
of exports we multiply these two matrices with a GN x GN matrix whose diagonal we fill with each industry's exports, $E$,
such that the basic equation behind the source decomposition is given by $V(I-A)^{-1}E$. In a simple example
with two countries (\textit{k} and \textit{l}) and industries (\textit{i} and \textit{j}) we can zoom in to see
the matrices' content:

\begin{align}
\begin{split}
V (I- A &)^{-1} E=
\begin{pmatrix}
v_{k}^{i}& 0& 0& 0\\
0& v_{k}^{j}& 0& 0\\
0& 0& v_{l}^{i}& 0\\
0& 0& 0& v_{l}^{j}
\end{pmatrix}
*
\begin{pmatrix}
b_{kk}^{ii}& b_{kk}^{ij}& b_{kl}^{ii}& b_{kl}^{ij}\\
b_{kk}^{ji}& b_{kk}^{jj}& b_{kl}^{ji}& b_{kl}^{jj}\\
b_{lk}^{ii}& b_{lk}^{ij}& b_{ll}^{ii}& b_{ll}^{ij}\\
b_{lk}^{ji}& b_{lk}^{jj}& b_{ll}^{ji}& b_{ll}^{jj}
\end{pmatrix}
*
\begin{pmatrix}
e_{k}^{i}& 0& 0& 0\\
0& e_{k}^{j}& 0& 0\\
0& 0& e_{l}^{i}& 0\\
0& 0& 0& e_{l}^{j}
\end{pmatrix}
=
\\
&
\begin{pmatrix}
v_{k}^{i}b_{kk}^{ii}e_{k}^{i}& v_{k}^{i} b_{kk}^{ij}e_{k}^{j}& v_{k}^{i}b_{kl}^{ii}e_{l}^{i}& v_{k}^{i}b_{kl}^{ij}e_{l}^{j}\\
v_{k}^{j}b_{kk}^{ji}e_{k}^{i}& v_{k}^{j}b_{kk}^{jj}e_{k}^{j}& v_{k}^{j}b_{kl}^{ji}e_{l}^{i}& v_{k}^{j}b_{kl}^{jj}e_{l}^{j}\\
v_{l}^{i}b_{lk}^{ii}e_{k}^{i}& v_{l}^{i}b_{lk}^{ij}e_{k}^{j}& v_{l}^{i}b_{ll}^{ii}e_{l}^{i}& v_{l}^{i}b_{ll}^{ij}e_{l}^{j}\\
v_{l}^{j}b_{lk}^{ji}e_{k}^{i}& v_{l}^{j}b_{lk}^{jj}e_{k}^{j}& v_{l}^{j}b_{ll}^{ji}e_{l}^{i}& v_{l}^{j}b_{ll}^{jj}e_{l}^{j}
\end{pmatrix}
=
\begin{pmatrix}
vae_{kk}^{ii}& vae_{kk}^{ij}& vae_{kl}^{ii}& vae_{kl}^{ij}\\
vae_{kk}^{ji}& vae_{kk}^{jj}& vae_{kl}^{ji}& vae_{kl}^{jj}\\
vae_{lk}^{ii}& vae_{lk}^{ij}& vae_{ll}^{ii}& vae_{ll}^{ij}\\
vae_{lk}^{ji}& vae_{lk}^{jj}& vae_{ll}^{ji}& vae_{ll}^{jj}
\end{pmatrix}
\end{split}
\end{align}
\begin{align*}
&v_{c}^{s} = \frac{va_{c}^{s}}{y_{c}^{s}} = 1 - a_{kc}^{is} - a_{kc}^{js} - a_{lc}^{js} - a_{lc}^{is} \hspace{2cm} (c \in k,l \hspace{.5cm} s \in i,j),\\
&\begin{pmatrix}
b_{kk}^{ii}& b_{kk}^{ij}& b_{kl}^{ii}& b_{kl}^{ij}\\
b_{kk}^{ji}& b_{kk}^{jj}& b_{kl}^{ji}& b_{kl}^{jj}\\
b_{lk}^{ii}& b_{lk}^{ij}& b_{ll}^{ii}& b_{ll}^{ij}\\
b_{lk}^{ji}& b_{lk}^{jj}& b_{ll}^{ji}& b_{ll}^{jj}
\end{pmatrix}
=
\begin{pmatrix}
1-a_{kk}^{ii}& -a_{kk}^{ij}& -a_{kl}^{ii}& -a_{kl}^{ij}\\
-a_{kk}^{ji}& 1-a_{kk}^{jj}& -a_{kl}^{ji}& -a_{kl}^{jj}\\
-a_{lk}^{ii}& -a_{lk}^{ij}& 1-a_{ll}^{ii}& -a_{ll}^{ij}\\
-a_{lk}^{ji}& -a_{lk}^{jj}& -a_{ll}^{ji}& 1-a_{ll}^{jj}
\end{pmatrix}^{-1},
\end{align*}
and
\begin{equation*}
a_{cf}^{su} = \frac{inp_{cf}^{su}}{y_{f}^{u}}  \hspace{2cm} (c,f \in k,l \hspace{.5cm} s,u \in i,j),
\end{equation*}
where \(v_{s}^{c}\) gives the share of industry \emph{s}'s value added,
\(va_{c}^{s},\) in output, \(y_{s}^{c}\), and \(e_{k}^{i}\)
indicates gross exports. Finally, \(a_{su}^{cf}\) denotes
the share of inputs, \(inp_{su}^{cf}\), in output. 
The elements of the $V(I-A)^{-1}E$ or $vae$ matrix are our estimates for the country-industry level value added
origins of each country-industry's exports. \textit{decompr} implements this algorithm into R to automate the process of
deriving the matrix. Equipped with it, researchers can calculate standard GVC indicators. 
Examples include \citet{dahuetal01}'s Vertical Specialisation ratio at the industry-level which can be derived
by summing for each country and industry across the value added of all foreign countries and industries.
Alternatively, the four dimensions of the matrix (source country, source industry, using country, using industry)
allow for industry-level gravity-type estimations of value added trade flows.

\subsection{Implentation}
The first step is to load the data and create the elements, there are all stored in a list (of class \textbf{decompr}).

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlstd{decompr_object} \hlkwb{<-} \hlkwd{load_tables_vectors}\hlstd{(} \hlkwc{x} \hlstd{= inter,}
                                       \hlkwc{y} \hlstd{= final,}
                                       \hlkwc{k} \hlstd{= countries,}
                                       \hlkwc{i} \hlstd{= industries,}
                                       \hlkwc{o} \hlstd{= out        )}
\end{alltt}
\end{kframe}
\end{knitrout}

We can now decompose the elements using the Leontief decomposition.
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlstd{lt} \hlkwb{<-} \hlkwd{leontief}\hlstd{( decompr_object )}
\end{alltt}
\end{kframe}
\end{knitrout}

In addition, a wrapper function called \textbf{decomp} is provided which performs both steps at once.
Though it is recomended that the atomic functions be used for large data sets.

@@

>>>>

The first step of the analytic process is to load the input object and create a \textbf{decompr} class object which contains the data structures for the decompositions. This step is not need when using the \textbf{decomp} wrapper function, more on this later.

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcom{# create the decompr object}
\hlstd{decompr_object} \hlkwb{<-} \hlkwd{load_tables_vectors}\hlstd{(} \hlkwc{x} \hlstd{= inter,}
                                       \hlkwc{y} \hlstd{= final,}
                                       \hlkwc{k} \hlstd{= countries,}
                                       \hlkwc{i} \hlstd{= industries,}
                                       \hlkwc{o} \hlstd{= out        )}

\hlcom{# inspect the content of the decompr object}
\hlkwd{ls}\hlstd{(decompr_object)}
\end{alltt}
\begin{verbatim}
##  [1] "A"         "Ad"        "Am"        "B"         "Bd"       
##  [6] "bigrownam" "Bm"        "E"         "Efd"       "Eint"     
## [11] "ESR"       "Exp"       "G"         "GN"        "k"        
## [16] "L"         "N"         "rownam"    "tot"       "Vc"       
## [21] "Vhat"      "X"         "Y"         "Yd"        "Ym"       
## [26] "z"         "z01"       "z02"       "z1"        "z2"
\end{verbatim}
\end{kframe}
\end{knitrout}

As can be seen above, a \textbf{decompr} class object is in fact a list containing thirty different objects. Depending on the choice of decomposition, all or some of these objects are used.

@@

\begin{equation}
VB = V (I-A)^{-1},
\end{equation}
where \textbf{V} is a N x N matrix with the diagonal representing the direct value added contribution of the industries,
\textbf{A} is the Input-Output coefficient matrix with dimension N x N, i.e. it gives the direct input flows between
industries required for 1\$ of output, $B = (I-A)^{-1}$ is then the so called Leontief inverse, and \textbf{T} indicates
a matrix transpose operation. \textbf{VB} gives thus a N x N matrix of so called value added multipliers,
which denote the amount of value added that the production of an industry's 1\$ of output or exports brings about
in all other industries. Looking from the perspective of the supplying industries, the matrix gives the value added that
they contribute to the using industry's production. If we multiply it with a N x N matrix whose diagonal specifies each
industry's total output, we get value added origins as absolute values instead of shares.

The application of the Leontief insight to ICIOs as opposed to national Input-Output tables for our source decomposition
is straightforward. \textbf{V} refers now to a vector of direct value added contribution of all industries across
the different countries. Its dimension is correspondingly 1 x GN. \textbf{A} is now of dimension GN x GN
and gives the industry flows including cross border relationships. Since we are interested in the value added origins
of exports we multiply these two matrices with a GN x GN matrix whose diagonal we fill with each industry's exports, $E$,
such that the basic equation behind the source decomposition is given by $V(I-A)^{-1}E$. In a simple example
with two countries (\textbf{k} and \textbf{l}) and industries (\textbf{i} and \textbf{j}) we can zoom in to see
the matrices' content:

\begin{align}
\begin{split}
V (I- A &)^{-1} E=
\begin{pmatrix}
v_{k}^{i}& 0& 0& 0\\
0& v_{k}^{j}& 0& 0\\
0& 0& v_{l}^{i}& 0\\
0& 0& 0& v_{l}^{j}
\end{pmatrix}
*
\begin{pmatrix}
b_{kk}^{ii}& b_{kk}^{ij}& b_{kl}^{ii}& b_{kl}^{ij}\\
b_{kk}^{ji}& b_{kk}^{jj}& b_{kl}^{ji}& b_{kl}^{jj}\\
b_{lk}^{ii}& b_{lk}^{ij}& b_{ll}^{ii}& b_{ll}^{ij}\\
b_{lk}^{ji}& b_{lk}^{jj}& b_{ll}^{ji}& b_{ll}^{jj}
\end{pmatrix}
*
\begin{pmatrix}
e_{k}^{i}& 0& 0& 0\\
0& e_{k}^{j}& 0& 0\\
0& 0& e_{l}^{i}& 0\\
0& 0& 0& e_{l}^{j}
\end{pmatrix}
=
\\
&
\begin{pmatrix}
v_{k}^{i}b_{kk}^{ii}e_{k}^{i}& v_{k}^{i} b_{kk}^{ij}e_{k}^{j}& v_{k}^{i}b_{kl}^{ii}e_{l}^{i}& v_{k}^{i}b_{kl}^{ij}e_{l}^{j}\\
v_{k}^{j}b_{kk}^{ji}e_{k}^{i}& v_{k}^{j}b_{kk}^{jj}e_{k}^{j}& v_{k}^{j}b_{kl}^{ji}e_{l}^{i}& v_{k}^{j}b_{kl}^{jj}e_{l}^{j}\\
v_{l}^{i}b_{lk}^{ii}e_{k}^{i}& v_{l}^{i}b_{lk}^{ij}e_{k}^{j}& v_{l}^{i}b_{ll}^{ii}e_{l}^{i}& v_{l}^{i}b_{ll}^{ij}e_{l}^{j}\\
v_{l}^{j}b_{lk}^{ji}e_{k}^{i}& v_{l}^{j}b_{lk}^{jj}e_{k}^{j}& v_{l}^{j}b_{ll}^{ji}e_{l}^{i}& v_{l}^{j}b_{ll}^{jj}e_{l}^{j}
\end{pmatrix}
=
\begin{pmatrix}
vae_{kk}^{ii}& vae_{kk}^{ij}& vae_{kl}^{ii}& vae_{kl}^{ij}\\
vae_{kk}^{ji}& vae_{kk}^{jj}& vae_{kl}^{ji}& vae_{kl}^{jj}\\
vae_{lk}^{ii}& vae_{lk}^{ij}& vae_{ll}^{ii}& vae_{ll}^{ij}\\
vae_{lk}^{ji}& vae_{lk}^{jj}& vae_{ll}^{ji}& vae_{ll}^{jj}
\end{pmatrix}
\end{split}
\end{align}
\begin{align*}
&v_{c}^{s} = \frac{va_{c}^{s}}{y_{c}^{s}} = 1 - a_{kc}^{is} - a_{kc}^{js} - a_{lc}^{js} - a_{lc}^{is} \hspace{2cm} (c \in k,l \hspace{.5cm} s \in i,j),\\
&\begin{pmatrix}
b_{kk}^{ii}& b_{kk}^{ij}& b_{kl}^{ii}& b_{kl}^{ij}\\
b_{kk}^{ji}& b_{kk}^{jj}& b_{kl}^{ji}& b_{kl}^{jj}\\
b_{lk}^{ii}& b_{lk}^{ij}& b_{ll}^{ii}& b_{ll}^{ij}\\
b_{lk}^{ji}& b_{lk}^{jj}& b_{ll}^{ji}& b_{ll}^{jj}
\end{pmatrix}
=
\begin{pmatrix}
1-a_{kk}^{ii}& -a_{kk}^{ij}& -a_{kl}^{ii}& -a_{kl}^{ij}\\
-a_{kk}^{ji}& 1-a_{kk}^{jj}& -a_{kl}^{ji}& -a_{kl}^{jj}\\
-a_{lk}^{ii}& -a_{lk}^{ij}& 1-a_{ll}^{ii}& -a_{ll}^{ij}\\
-a_{lk}^{ji}& -a_{lk}^{jj}& -a_{ll}^{ji}& 1-a_{ll}^{jj}
\end{pmatrix}^{-1},
\end{align*}
and
\begin{equation*}
a_{cf}^{su} = \frac{inp_{cf}^{su}}{y_{f}^{u}}  \hspace{2cm} (c,f \in k,l \hspace{.5cm} s,u \in i,j),
\end{equation*}
where \(v_{s}^{c}\) gives the share of industry \emph{s}'s value added,
\(va_{c}^{s},\) in output, \(y_{s}^{c}\), and \(e_{k}^{i}\)
indicates gross exports. Finally, \(a_{su}^{cf}\) denotes
the share of inputs, \(inp_{su}^{cf}\), in output. 
The elements of the $V(I-A)^{-1}E$ or $vae$ matrix are our estimates for the country-industry level value added
origins of each country-industry's exports. \textbf{decompr} implements this algorithm into R to automate the process of
deriving the matrix. Equipped with it, researchers can calculate standard GVC indicators. 
Examples include \citet{dahuetal01}'s Vertical Specialisation ratio at the industry-level which can be derived
by summing for each country and industry across the value added of all foreign countries and industries.
Alternatively, the four dimensions of the matrix (source country, source industry, using country, using industry)
allow for industry-level gravity-type estimations of value added trade flows.

\subsection{Implentation}
As described, in \cref{sec:data}, the first step of our analytic process is to construct a \textbf{decompr} object using the \textbf{load\_tables\_vectors} function. After this, we can use the \textbf{leontief} function to apply to Leontief decomposition.
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlstd{lt} \hlkwb{<-} \hlkwd{leontief}\hlstd{( decompr_object )}
\end{alltt}
\end{kframe}
\end{knitrout}

In addition, a wrapper function called \textbf{decomp} is provided which integrates both elements of the workflow into a single function.
We recomended that the atomic functions be used for large data sets, however, for small data sets this is an easy way to derive the results immediatly.
The \textbf{decomp} function requires a method to be specified (see \textbf{help("decomp) for details}), is none is provided, the function will default to \textbf{leontief}.

@@

Furthermore, it is also possible to derive the results of the Wang-Wei-Zhu decomposition directly, using the \textbf{decomp} function.

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlstd{w2} \hlkwb{<-}  \hlkwd{decomp}\hlstd{(} \hlkwc{x} \hlstd{= inter,}
               \hlkwc{y} \hlstd{= final,}
               \hlkwc{k} \hlstd{= countries,}
               \hlkwc{i} \hlstd{= industries,}
               \hlkwc{o} \hlstd{= out,}
               \hlkwc{method} \hlstd{=} \hlstr{"wwz"} \hlstd{)}
\end{alltt}
\end{kframe}
\end{knitrout}

Both these processes will yield the same results.
